---
title: "Hypertensive emergency meta-analysis"
author: "Joonatan Palmu"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output:
    rmdformats::html_clean:
        self_contained: true
        thumbnails: false
        lightbox: true
        gallery: true
---

# Init

```{r options, echo = FALSE}
knitr::opts_chunk$set(include = TRUE, echo = TRUE, message = FALSE, cache=FALSE, warning=FALSE)
```

## Libraries

```{css, echo = FALSE}
.main-container {
  max-width:1920px !important;
}

table {
  width: 100% !important;
}

th:not(:first-child) {
  text-align:right !important;
}
```

```{r, echo = FALSE}
library(magrittr)
```

## Ggplot2

```{r}
{ ggthemes::theme_tufte(base_family = "sans", base_size = 12) +
    ggplot2::theme(axis.text = ggplot2::element_text(colour = "black", size = 10),
                   panel.border = ggplot2::element_rect(colour = "black", fill = NA),
                   panel.background = ggplot2::element_rect(fill = "white", colour = "white"),
                   plot.background = ggplot2::element_rect(fill = "white", colour = "white")) } %>%
  ggplot2::theme_set()
```

```


# Data definitions

## Systolic BP

```{r, echo = FALSE}
df_bp <- tibble::tibble(study = c("Shin et al. (2021)",
                                  "Reis et al. (2020)",
                                  "Lane et al. (2009)"),
                        HR = c(1.01,
                               1.42^0.1,
                               1.001),
                        conf_low = c(1.01,
                                     1.24^0.1,
                                     0.996),
                        conf_high = c(1.02,
                                      1.62^0.1,
                                      1.006)) %>%
  dplyr::mutate(dplyr::across(c(HR, conf_low, conf_high), ~.^10)) %>% 
  dplyr::mutate(SE = (conf_high - conf_low) / (2 * 1.96),
                logHR = log(HR))
```

```{r, echo = FALSE}
df_bp %>% knitr::kable()
```

## Age

```{r, echo = FALSE}
df_age <- tibble::tibble(study = c("Afonso et al. (2011)",
                                   "Lane et al. (2009)",
                                   "Yizengaw et al. (2022)"),
                         HR = c(1.025,
                                1.04,
                                1.016),
                         conf_low = c(1.016,
                                      1.03,
                                      0.969),
                         conf_high = c(1.034,
                                       1.05,
                                       1.065)) %>%
  dplyr::mutate(dplyr::across(c(HR, conf_low, conf_high), ~.^10)) %>% 
  dplyr::mutate(SE = (conf_high - conf_low) / (2 * 1.96),
                logHR = log(HR))
```

```{r, echo = FALSE}
df_age %>% knitr::kable()
```

## Male

```{r, echo = FALSE}
df_sex <- tibble::tibble(study = c("Shin et al. (2021)",
                                   "Reis et al. (2020)",
                                   "Lane el al. (2009)"),
                         HR = c(1.54,
                                1/0.97,
                                0.95),
                         conf_low = c(1.22,
                                      1/1.72,
                                      0.65),
                         conf_high = c(1.94,
                                       1/0.55,
                                       1.41)) %>%
  dplyr::mutate(SE = (conf_high - conf_low) / (2 * 1.96),
                logHR = log(HR))
```

```{r, echo = FALSE}
df_sex %>% knitr::kable()
```


# Random-effects meta-analysis

## Systolic BP

```{r}
metafor::rma(yi = logHR, sei = SE, data = df_bp, method = "REML") %>%
  summary()
```

## Age

```{r}
metafor::rma(yi = logHR, sei = SE, data = df_age, method = "REML") %>%
  summary()
```

## Men

```{r}
metafor::rma(yi = logHR, sei = SE, data = df_sex, method = "REML") %>%
  summary()
```

# Forest plot

```{r}
make_plot <- function(df, xlab = "HR (95% CI)") {
  df_meta <- metafor::rma(yi = logHR, sei = SE, data = df, method = "REML") %>%
    broom::tidy(conf.int = TRUE) %>%
    dplyr::select(HR = estimate, conf_low = conf.low, conf_high = conf.high) %>%
    dplyr::mutate(dplyr::across(c(HR, conf_low, conf_high), exp)) %>%
    dplyr::mutate(study = "Overall")

  df_plot <- dplyr::bind_rows(df, df_meta) %>% 
    dplyr::mutate_if(is.numeric, round, 2) %>%
    dplyr::mutate(label = glue::glue("{format(HR, 2)} ({format(conf_low, 2)}-{format(conf_high, 2)})")) %>%
    dplyr::mutate(study = stringr::str_replace_all(study, ".et.al..", " et al. ")) 

  forestplot::forestplot(labeltext = cbind(df_plot$study %>% as.character(), df_plot$label),
                         mean = df_plot$HR,
                         lower = df_plot$conf_low,
                         upper = df_plot$conf_high,
                         align = c("l", "l", "l"),
                         graph.pos = 2,
                         title = "",
                         xlog = FALSE,
                         xlab = xlab,
                         is.summary = c(rep(FALSE, nrow(df_plot)-1), TRUE),
                         txt_gp = forestplot::fpTxtGp(label = grid::gpar(cex = 1.25),
                                                      ticks = grid::gpar(cex = 1.1),
                                                      xlab = grid::gpar(cex = 1.2),
                                                      title = grid::gpar(cex = 1.2)),
                         xticks = seq(0.5, 2.0, 0.5),
                         clip =c(0.9, 1.5),
                         zero = 1,
                         lineheight = grid::unit(10, "mm"),
                         boxsize = 0.15,
                         colgap = grid::unit(4, "mm"),
                         lwd.ci = 1) %>%
    print() %>% 
    grid::grid.grabExpr()
}
```

```{r}
cowplot::plot_grid(make_plot(df_age, xlab = "HR (95% CI) per 10-year increase in age"),
                   make_plot(df_bp, xlab = "HR (95% CI) per 10 mmHg increase in blood pressure"),
                   make_plot(df_sex),
                   ncol = 1,
                   hjust = 0,
                   labels = c("A. Age",
                              "B. Systolic blood pressure",
                              "C. Male sex")) %>%
  ggplot2::ggsave("cache/forest.png",
                  plot = .,
                  width = 6,
                  height = 10,
                  unit = "in",
                  dpi = 300)
```

<img src = "cache/forest.png" />

# Incidence

```{r}
df_incidence <- readr::read_tsv("incidence.tsv") %>%
    dplyr::arrange(Study)
```

## Study rates

```{r}
myrate <- function(df, outcome) {
  meta_result <- df %>%
    dplyr::rename(N = !!glue::glue("{outcome}_N"),
                  events = !!glue::glue("{outcome}_deaths")) %>% 
    dplyr::select(Study, followup, N, events) %>%
    dplyr::mutate(person_time = followup*N) %>%
    tidyr::drop_na() %>% 
    meta::metarate(event = events, time = person_time, studlab = Study, data = .)

  tibble::tibble(study = meta_result$studlab,
                 event = meta_result$event,
                 time = meta_result$time,
                 rate = meta_result$TE) %>%
    dplyr::mutate(dplyr::across(c(rate), exp))
}
```

```{r}
df_rate <- c("HTNC", "HTNU", "HTNE") %>%
  setNames(., .) %>%
  purrr::map_df(~myrate(df_incidence, .x), .id = "type") %>%
  dplyr::mutate(lower_ci = rate, upper_ci = rate)
```

## Meta-analysis for incidence rates


```{r}
mymetarate <- function(df, outcome) {
  meta_result <- df %>%
    dplyr::rename(N = !!glue::glue("{outcome}_N"),
                  events = !!glue::glue("{outcome}_deaths")) %>% 
    dplyr::select(Study, followup, N, events) %>%
    dplyr::mutate(person_time = followup*N) %>%
    tidyr::drop_na() %>% 
    meta::metarate(event = events, time = person_time, studlab = Study, data = .)
  
  tibble::tibble(method = c("Fixed effect", "Random effects"),
                 rate = c(meta_result$TE.common, meta_result$TE.random),
                 se = c(meta_result$seTE.fixed, meta_result$seTE.random),
                 lower_ci = c(meta_result$lower.common, meta_result$lower.random),
                 upper_ci = c(meta_result$upper.common, meta_result$upper.random),
                 I2 = rep(meta_result$I2, 2)) %>%
    dplyr::mutate(dplyr::across(c(rate, lower_ci, upper_ci), exp))
}

```

```{r}
df_metarate_short <- c("HTNC", "HTNU", "HTNE") %>%
  setNames(., .) %>%
  purrr::map_df(~mymetarate(df_incidence %>% dplyr::filter(followup <= 1), .x), .id = "type") %>%
  dplyr::filter(method == "Random effects") %>%
  dplyr::select(-method) %>%
  dplyr::mutate(study = "Overall")
```

```{r}
df_metarate_short %>% knitr::kable()
```

```{r}
df_metarate_long <- c("HTNC", "HTNU", "HTNE") %>%
  setNames(., .) %>%
  purrr::map_df(~mymetarate(df_incidence %>% dplyr::filter(followup > 1), .x), .id = "type") %>%
  dplyr::filter(method == "Random effects") %>%
  dplyr::select(-method) %>%
  dplyr::mutate(study = "Overall")
```

```{r}
df_metarate_long %>% knitr::kable()
```


## Compare incidence differences

```{r}
ztest <- function(event_a, event_b) {
  rate_diff <- abs(event_a[["rate"]] - event_b[["rate"]])
  se_diff <- sqrt(event_a[["se"]]^2 + event_b[["se"]]^2)
  z_stat <- rate_diff / se_diff
  2 * (1 - pnorm(z_stat))
}
```

```{r}
c_htnc_short <- df_metarate_short %>% dplyr::filter(type == "HTNC") %>% dplyr::select(rate, se)
c_htnc_long <- df_metarate_long %>% dplyr::filter(type == "HTNC") %>% dplyr::select(rate, se)
c_htnu_long <- df_metarate_long %>% dplyr::filter(type == "HTNU") %>% dplyr::select(rate, se)
c_htne_long <- df_metarate_long %>% dplyr::filter(type == "HTNE") %>% dplyr::select(rate, se)
```

```{r}
ztest(c_htnc_short, c_htne)
```


```{r}
ztest(c_htnc, c_htne)
```

```{r}
ztest(c_htnc, c_htne)
```

```{r}
ztest(c_htnu, c_htne)
```


## Plot

```{r}
incplot <- function(df_plot, lab = "Annual mortality rate (95% CI)", clip = c(0, 0.4)) {
  forestplot::forestplot(labeltext = rbind(c("Study", "Incidence", "Events", "Person time"),
                                           cbind(df_plot$study %>% as.character(),
                                                 df_plot$label,
                                                 ifelse(is.na(df_plot$event), "", sprintf("%4d", df_plot$event)),
                                                 ifelse(is.na(df_plot$time), "", sprintf("%.1f", df_plot$time)))),
                         mean = c(NA, df_plot$rate),
                         lower = c(NA, df_plot$lower_ci),
                         upper = c(NA, df_plot$upper_ci),
                         align = c("l", "l", "r", "r"),
                         graph.pos = 3,
                         title = "",
                         xlog = FALSE,
                         xlab = lab,
                         is.summary = c(rep(FALSE, nrow(df_plot)), TRUE),
                         txt_gp = forestplot::fpTxtGp(label = grid::gpar(cex = 1.25),
                                                      ticks = grid::gpar(cex = 1.1),
                                                      xlab = grid::gpar(cex = 1.2),
                                                      title = grid::gpar(cex = 1.2)),
                         graphwidth = grid::unit(4, "cm"),
                         xticks = seq(0, max(clip), 0.1),
                         clip = clip,
                         zero = 0,
                         lineheight = grid::unit(10, "mm"),
                         boxsize = 0.15,
                         colgap = grid::unit(4, "mm"),
                         lwd.ci = 1) %>%
    print() %>% 
    grid::grid.grabExpr()
}
```

```{r}
dplyr::bind_rows(df_rate, df_metarate) %>%
  dplyr::filter(type == "HTNC") %>%
  dplyr::mutate(label = dplyr::case_when(study != "Overall" ~ glue::glue("{sprintf('%.2f', rate)}"),
                                         TRUE ~ glue::glue("{sprintf('%.2f', rate)} ({sprintf('%.2f', lower_ci)}-{sprintf('%.2f', upper_ci)})"))) %>% 
  incplot(clip = c(0, 0.4)) %>%
  ggplot2::ggsave("cache/incidence_htnc.png",
                  plot = .,
                  width = 10,
                  height = 9,
                  unit = "in",
                  dpi = 300)
```


<img src = "cache/incidence_htnc.png" />

```{r}
dplyr::bind_rows(df_rate, df_metarate) %>%
  dplyr::filter(type == "HTNU") %>%
  dplyr::mutate(label = dplyr::case_when(study != "Overall" ~ glue::glue("{sprintf('%.2f', rate)}"),
                                         TRUE ~ glue::glue("{sprintf('%.2f', rate)} ({sprintf('%.2f', lower_ci)}-{sprintf('%.2f', upper_ci)})"))) %>% 
  incplot(clip = c(0, 0.4)) %>%
  ggplot2::ggsave("cache/incidence_htnu.png",
                  plot = .,
                  width = 10,
                  height = 3,
                  unit = "in",
                  dpi = 300)
```

<img src = "cache/incidence_htnu.png" />

```{r}
dplyr::bind_rows(df_rate, df_metarate) %>%
  dplyr::filter(type == "HTNE") %>%
  dplyr::mutate(label = dplyr::case_when(study != "Overall" ~ glue::glue("{sprintf('%.2f', rate)}"),
                                         TRUE ~ glue::glue("{sprintf('%.2f', rate)} ({sprintf('%.2f', lower_ci)}-{sprintf('%.2f', upper_ci)})"))) %>% 
  incplot(clip = c(0, 0.4)) %>%
  ggplot2::ggsave("cache/incidence_htne.png",
                  plot = .,
                  width = 10,
                  height = 4,
                  unit = "in",
                  dpi = 300)
```

<img src = "cache/incidence_htne.png" />
