---
title: "Hypertensive emergency meta-analysis"
author: "Joonatan Palmu"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output:
    rmdformats::html_clean:
        self_contained: true
        thumbnails: false
        lightbox: true
        gallery: true
---

# Init

```{r options, echo = FALSE}
knitr::opts_chunk$set(include = TRUE, echo = TRUE, message = FALSE, cache=FALSE, warning=FALSE)
```

## Libraries

```{css, echo = FALSE}
.main-container {
  max-width:1920px !important;
}

table {
  width: 100% !important;
}

th:not(:first-child) {
  text-align:right !important;
}
```

```{r, echo = FALSE}
library(magrittr)
```

## Ggplot2

```{r}
{ ggthemes::theme_tufte(base_family = "sans", base_size = 12) +
    ggplot2::theme(axis.text = ggplot2::element_text(colour = "black", size = 10),
                   panel.border = ggplot2::element_rect(colour = "black", fill = NA),
                   panel.background = ggplot2::element_rect(fill = "white", colour = "white"),
                   plot.background = ggplot2::element_rect(fill = "white", colour = "white")) } %>%
  ggplot2::theme_set()
```

```


# Data definitions

## Systolic BP

```{r, echo = FALSE}
df_bp <- tibble::tibble(study = c("Shin et al. (2021)",
                                  "Reis et al. (2020)",
                                  "Lane et al. (2009)"),
                        HR = c(1.01,
                               1.42^0.1,
                               1.001),
                        conf_low = c(1.01,
                                     1.24^0.1,
                                     0.996),
                        conf_high = c(1.02,
                                      1.62^0.1,
                                      1.006)) %>%
  dplyr::mutate(dplyr::across(c(HR, conf_low, conf_high), ~.^10)) %>% 
  dplyr::mutate(SE = (conf_high - conf_low) / (2 * 1.96),
                logHR = log(HR))
```

```{r, echo = FALSE}
df_bp %>% knitr::kable()
```

## Age

```{r, echo = FALSE}
df_age <- tibble::tibble(study = c("Afonso et al. (2011)",
                                   "Lane et al. (2009)",
                                   "Yizengaw et al. (2022)"),
                         HR = c(1.025,
                                1.04,
                                1.016),
                         conf_low = c(1.016,
                                      1.03,
                                      0.969),
                         conf_high = c(1.034,
                                       1.05,
                                       1.065)) %>%
  dplyr::mutate(dplyr::across(c(HR, conf_low, conf_high), ~.^10)) %>% 
  dplyr::mutate(SE = (conf_high - conf_low) / (2 * 1.96),
                logHR = log(HR))
```

```{r, echo = FALSE}
df_age %>% knitr::kable()
```

## Male

```{r, echo = FALSE}
df_sex <- tibble::tibble(study = c("Shin et al. (2021)",
                                   "Reis et al. (2020)",
                                   "Lane el al. (2009)"),
                         HR = c(1.54,
                                1/0.97,
                                0.95),
                         conf_low = c(1.22,
                                      1/1.72,
                                      0.65),
                         conf_high = c(1.94,
                                       1/0.55,
                                       1.41)) %>%
  dplyr::mutate(SE = (conf_high - conf_low) / (2 * 1.96),
                logHR = log(HR))
```

```{r, echo = FALSE}
df_sex %>% knitr::kable()
```


# Random-effects meta-analysis

## Systolic BP

```{r}
metafor::rma(yi = logHR, sei = SE, data = df_bp, method = "REML") %>%
  summary()
```

## Age

```{r}
metafor::rma(yi = logHR, sei = SE, data = df_age, method = "REML") %>%
  summary()
```

## Men

```{r}
metafor::rma(yi = logHR, sei = SE, data = df_sex, method = "REML") %>%
  summary()
```

# Forest plot

```{r}
make_plot <- function(df, xlab = "HR (95% CI)") {
  df_meta <- metafor::rma(yi = logHR, sei = SE, data = df, method = "REML") %>%
    broom::tidy(conf.int = TRUE) %>%
    dplyr::select(HR = estimate, conf_low = conf.low, conf_high = conf.high) %>%
    dplyr::mutate(dplyr::across(c(HR, conf_low, conf_high), exp)) %>%
    dplyr::mutate(study = "Overall")

  df_plot <- dplyr::bind_rows(df, df_meta) %>% 
    dplyr::mutate_if(is.numeric, round, 2) %>%
    dplyr::mutate(label = glue::glue("{format(HR, 2)} ({format(conf_low, 2)}-{format(conf_high, 2)})")) %>%
    dplyr::mutate(study = stringr::str_replace_all(study, ".et.al..", " et al. ")) 

  forestplot::forestplot(labeltext = cbind(df_plot$study %>% as.character(), df_plot$label),
                         mean = df_plot$HR,
                         lower = df_plot$conf_low,
                         upper = df_plot$conf_high,
                         align = c("l", "l", "l"),
                         graph.pos = 2,
                         title = "",
                         xlog = FALSE,
                         xlab = xlab,
                         is.summary = c(rep(FALSE, nrow(df_plot)-1), TRUE),
                         txt_gp = forestplot::fpTxtGp(label = grid::gpar(cex = 1.25),
                                                      ticks = grid::gpar(cex = 1.1),
                                                      xlab = grid::gpar(cex = 1.2),
                                                      title = grid::gpar(cex = 1.2)),
                         xticks = seq(0.5, 2.0, 0.5),
                         clip =c(0.9, 1.5),
                         zero = 1,
                         lineheight = grid::unit(10, "mm"),
                         boxsize = 0.15,
                         colgap = grid::unit(4, "mm"),
                         lwd.ci = 1) %>%
    print() %>% 
    grid::grid.grabExpr()
}
```

```{r}
cowplot::plot_grid(make_plot(df_age, xlab = "HR (95% CI) per 10-year increase in age"),
                   make_plot(df_bp, xlab = "HR (95% CI) per 10 mmHg increase in blood pressure"),
                   make_plot(df_sex),
                   ncol = 1,
                   hjust = 0,
                   labels = c("A. Age",
                              "B. Systolic blood pressure",
                              "C. Male sex")) %>%
  ggplot2::ggsave("cache/forest.png",
                  plot = .,
                  width = 6,
                  height = 10,
                  unit = "in",
                  dpi = 300)
```

<img src = "cache/forest.png" />

# Incidence

```{r}
df_incidence <- readr::read_tsv("incidence.tsv")
```

## Study rates

```{r}
myrate <- function(df, outcome) {
  meta_result <- df_incidence %>%
    dplyr::rename(N = !!glue::glue("{outcome}_N"),
                  events = !!glue::glue("{outcome}_deaths")) %>% 
    dplyr::select(Study, followup, N, events) %>%
    dplyr::mutate(person_time = followup*N) %>%
    tidyr::drop_na() %>% 
    meta::metarate(event = events, time = person_time, studlab = Study, data = .)

  tibble::tibble(rate = meta_result$TE) %>%
    dplyr::mutate(dplyr::across(c(rate), exp))
}
```

```{r}
df_rate <- c("HTNC", "HTNU", "HTNE") %>%
  setNames(., .) %>%
  purrr::map_df(~myrate(df_incidence, .x), .id = "type") %>%
  dplyr::mutate(type = factor(type, levels = c("HTNC", "HTNU", "HTNE"))) %>%
  dplyr::mutate(type = forcats::fct_recode(type,
                                           "Hypertensive crisis" = "HTNC",
                                           "Hypertensive urgency" = "HTNU",
                                           "Hypertensive emergency" = "HTNE"))
```

## Meta-analysis for incidence rates


```{r}
mymetarate <- function(df, outcome) {
  meta_result <- df_incidence %>%
    dplyr::rename(N = !!glue::glue("{outcome}_N"),
                  events = !!glue::glue("{outcome}_deaths")) %>% 
    dplyr::select(Study, followup, N, events) %>%
    dplyr::mutate(person_time = followup*N) %>%
    tidyr::drop_na() %>% 
    meta::metarate(event = events, time = person_time, studlab = Study, data = .)
  
  tibble::tibble(method = c("Fixed effect", "Random effects"),
                 rate = c(meta_result$TE.common, meta_result$TE.random),
                 lower_ci = c(meta_result$lower.common, meta_result$lower.random),
                 upper_ci = c(meta_result$upper.common, meta_result$upper.random),
                 I2 = rep(meta_result$I2, 2)) %>%
    dplyr::mutate(dplyr::across(c(rate, lower_ci, upper_ci), exp))
}

```

```{r}
df_metarate <- c("HTNC", "HTNU", "HTNE") %>%
  setNames(., .) %>%
  purrr::map_df(~mymetarate(df_incidence, .x), .id = "type") %>%
  dplyr::filter(method == "Random effects") %>%
  dplyr::select(-method) %>%
  dplyr::mutate(type = factor(type, levels = c("HTNC", "HTNU", "HTNE"))) %>%
  dplyr::mutate(type = forcats::fct_recode(type,
                                           "Hypertensive crisis" = "HTNC",
                                           "Hypertensive urgency" = "HTNU",
                                           "Hypertensive emergency" = "HTNE"))
```

```{r}
df_metarate %>% knitr::kable()
df_rate
```

## Plot

```{r}
{ ggplot2::ggplot(df_metarate, ggplot2::aes(x = forcats::fct_rev(type), y = rate)) +
    ggplot2::geom_jitter(data = df_rate, size = 4, width = 0.3, alpha = 0.4, stroke=NA) +
    ggplot2::geom_point(size = 8, shape = 18) +
    ggplot2::geom_errorbar(ggplot2::aes(ymin = lower_ci, ymax = upper_ci), width = 0.1, size = 0.5) +
    ggplot2::coord_flip() +
    ggplot2::scale_x_discrete(name = NULL, expand = c(0, 0.4)) +
    ggplot2::scale_y_continuous(name = "Annual mortality rate (95% CI)",
                                expand = c(0, 0),
                                breaks = seq(0, 1, 0.1),
                                limits = c(0, 0.42)) } %>%
  ggplot2::ggsave(filename = "cache/incidence.png", width = 8, height = 6, dpi = 300)
```

<img src = "cache/incidence.png" />
